\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-12-20 flux.tex
Fix typo.
Clarify notation.

% notation
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
\newcommand{\tree}{\tau}							% Phylogeny
\newcommand{\vn}{n}									% Number of viruses
\newcommand{\sn}{k}									% Number of sera
\setlength{\arraycolsep}{2pt}
\newcommand{\twomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
\newcommand{\fourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}		% pretty inline matrix

@@

Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
In the case of HI, the measurement of cross-reactivity takes the form of a titer $H_{ij}$ representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.  
These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.

@@

Here, we follow previous models in representing antigenic locations as points in an $N$-dimensional antigenic map. 
Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions. 
We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.

Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
Virus and serum may be isolated from / raised against the same strain and have different cartographic locations.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
This gives a set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || X_i - Y_j ||_2.
\end{equation}
With $H_{ij}$ as the log$_2$ titer of virus $i$ against serum $j$, immunological distance can be defined as
\begin{equation}
	d_{ij} =  S_j - H_{ij},
\end{equation}
where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} \label{hij}
	H_{ij} \sim \mbox{TruncatedNormal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing a particular titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

%%% hij_likelihood %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $S_j = \mathrm{log}_2 1280 = 10.32$.  
	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
	} 
	\label{hij_likelihood} 
\end{figure}

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.

We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, scaling of the entire set of virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ and scaling $\mdssd$.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.  
Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $S_j - \delta_{ij}$, but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

Additionally, we follow the same logic and assume that particular virus isolates are more reactive in general than other virus isolates.
With virus reactivity included, the expectation of $H_{ij}$ becomes $(V_i+S_j)/2 - \delta_{ij}$ and the vector of `virus effects' $V_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\mathbf{V}$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.

With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $S_j$ and virus effects $V_i$.

@@

As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution $\mathrm{Normal}(\mu,\Sigma)$ in which the expectation of the first antigenic dimension increases with date of virus sampling.
Here, $\mu = \twomatrix{\beta \, y_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
To do this, we incorporate a prior on virus locations according to the evolutionary process
\begin{equation}
	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
\end{equation}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \driftsd) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\driftsd,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}

@@

There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
The first issue lay in summarizing the posterior distribution of virus and serum locations.
Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift dramatically.
Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at $\{0,0\}$.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.

Running the fully parameterized model including the phylogenetic diffusion prior resulted a pathological outcome in which all virus and serum locations are compressed into a single point, diffusion volatility $\diffusionsd$ is tends toward zero and MDS variance $\mdssd^2$ tends towards infinity.
In this case, $p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd)$ is reduced, but $p(\mathbf{X} | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
To avoid this problematic outcome we conducted a two stage analysis.
We first estimated values of $\mdssd$, $\beta$ and $\driftsd$ using HI data alone, and then held these values constant in a subsequent analysis that included the phylogenetic diffusion prior.

@@

MCMC was used to sample virus locations $\mathbf{X}$, serum locations $\mathbf{Y}$, virus effects $\mathbf{V}$, serum effects $\mathbf{S}$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.

>>>>

% notation
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\virus}{\mathbf{x}}						% virus coordinate
\newcommand{\serum}{\mathbf{y}}						% serum coordinate
\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
\newcommand{\ve}{v}									% virus effect
\newcommand{\se}{s}									% serum effect
\newcommand{\ves}{\mathbf{v}}						% set of virus effects
\newcommand{\ses}{\mathbf{s}}						% set of serum effects
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
\newcommand{\tree}{\tau}							% Phylogeny
\newcommand{\vn}{n}									% Number of viruses
\newcommand{\sn}{k}									% Number of sera
\setlength{\arraycolsep}{2pt}
\newcommand{\twomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
\newcommand{\fourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}		% pretty inline matrix

@@

Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding of another virus.  
These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.

@@

Here, we follow previous models in representing antigenic locations as points in an $N$-dimensional antigenic map. 
Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions $P$. 
We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.

Let $\virus_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = \twomatrix{x_{i1}}{x_{i2}}$ for $P=2$. 
Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = \twomatrix{y_{j1}}{y_{j2}}$ for $P=2$.
This yields an $\vn \times P$ matrix $\viruses$ of virus coordinates and an $\sn \times P$ matrix $\sera$ of serum coordinates.
Virus and serum may be isolated from / raised against the same strain and have different cartographic locations. 
This gives a set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || \virus_i - \serum_j ||_2.
\end{equation}

Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
\begin{equation}
	d_{ij} =  \se_j - H_{ij},
\end{equation}
where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
In following multidimensional scaling (MDS), these approaches attempt to optimize over $\viruses$ and $\sera$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.

Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} \label{hij}
	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing an exact titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
The `serum effect' $\se_j$ represents the expected log$_2$ titer between virus $i$ and serum $j$ when $\delta_{ij}=0$ and is initially fixed such that $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.

HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

%%% hij_likelihood %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
	} 
	\label{hij_likelihood} 
\end{figure}

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior on virus and serum locations
\begin{eqnarray}
	\virus_i &\sim& \mathrm{Normal}(\mu,\Sigma) \nonumber \\
	\serum_j &\sim& \mathrm{Normal}(\mu,\Sigma),
\end{eqnarray}
with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.

We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
Metropolis-Hastings proposals include moves to individual virus and serum locations $\virus_i$ and $\serum_j$, scaling of the entire set of virus and serum locations $\viruses$ and $\sera$ and scaling $\mdssd$.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
However, this model may biased in some circumstances.
In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses$ is estimated rather than fixed.
We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.
The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

Additionally, we follow the same logic and assume that some virus isolates are more reactive in general than other virus isolates.
With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of `virus effects' $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.

With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $\se_j$ and virus effects $\ve_i$.

@@

As presented, virus and serum locations $\viruses$ and $\sera$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution in which the expectation of dimension 1 increases with date of virus sampling
\begin{equation}
	\virus_i \sim \mathrm{Normal}(\mu, \Sigma),
\end{equation}
where $\mu = \twomatrix{\beta \, t_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
To do this, we incorporate a prior on virus locations according to the evolutionary process
\begin{equation}
	(\virus_1,\ldots,\virus_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
\end{equation}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\viruses | \tree, \diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\viruses,\sera | \mathbf{H},\tree) = \frac{ p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
	p(\viruses | \beta, \driftsd) \;
	p(\viruses | \tree, \diffusionsd) \; 
	p(\sera, \ses, \ves, \mdssd, \beta, \driftsd, \diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}

@@

There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
The first issue lay in summarizing the posterior distribution of virus and serum locations.
Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift dramatically.
Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.

Running the fully parameterized model including the phylogenetic diffusion prior resulted a pathological outcome in which all virus and serum locations are compressed into a single point, diffusion volatility $\diffusionsd$ is tends toward zero and MDS variance $\mdssd^2$ tends towards infinity.
In this case, $p(\mathbf{H}|\viruses,\sera,\ses,\ves, \mdssd)$ is reduced, but $p(\viruses | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
To avoid this problematic outcome we conducted a two stage analysis.
We first estimated values of $\mdssd$, $\beta$ and $\driftsd$ using HI data alone, and then held these values constant in a subsequent analysis that included the phylogenetic diffusion prior.

@@

MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
